Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
re-add logging

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 50 imputations, 20 iterations (slow but thorough)
analysis_mode <- "test"  # Change to "final" for production run

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 2
} else if (analysis_mode == "final") {
  n_imputations <- 50
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

@@

ps_model_vars <- variable_lists$ps_model_vars
outcome_model_vars <- variable_lists$outcome_model_vars
```

@@

```{r mice-run}

@@

## Imputation Diagnostics

**Note:** In the stripplots below, the leftmost column (labeled 1) shows the **original observed data** with missing values. Columns 2 through `r n_imputations + 1` show the imputed datasets.

```{r imputation-diagnostics}
# Check imputed values for key variables
# In mice stripplots: .imp = 1 is original data, .imp = 2+ are imputations

# CCI - compare observed vs imputed
if(sum(is.na(df_mice$CCI)) > 0) {
  # Density plot: Blue = observed data, Red = imputed values
  densityplot(mice_results, ~ CCI, 
              main = "Charlson Comorbidity Index (CCI)") |> print()

  # Stripplot: Column 1 = original data, Columns 2+ = imputed datasets
  stripplot(mice_results, CCI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "CCI Distribution: Original vs Imputed") |> print()
}

# DCSI - compare observed vs imputed  
if(sum(is.na(df_mice$DCSI)) > 0) {
  densityplot(mice_results, ~ DCSI,
              main = "Diabetes Complications Severity Index (DCSI)") |> print()

  stripplot(mice_results, DCSI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "DCSI Distribution: Original vs Imputed") |> print()
}

# Baseline VA - compare observed vs imputed
if(sum(is.na(df_mice$baseline_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ baseline_VA_logMAR,
              main = "Baseline Visual Acuity (logMAR)") |> print()

  stripplot(mice_results, baseline_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Baseline VA Distribution: Original vs Imputed") |> print()
}

# Intermediate VA - compare observed vs imputed
if(sum(is.na(df_mice$intermediate_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ intermediate_VA_logMAR,
              main = "Intermediate Visual Acuity (logMAR)") |> print()

  stripplot(mice_results, intermediate_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Intermediate VA Distribution: Original vs Imputed") |> print()
}

# Outcome VA - compare observed vs imputed
if(sum(is.na(df_mice$outcome_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ outcome_VA_logMAR,
              main = "Outcome Visual Acuity at 2 Years (logMAR)") |> print()

  stripplot(mice_results, outcome_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "2-Year Outcome VA Distribution: Original vs Imputed") |> print()
}

# Person DR - compare observed vs imputed (categorical)
if(sum(is.na(df_mice$person_dr)) > 0) {
  # For categorical variables, check proportions across imputations
  # Extract each imputation and tabulate
  dr_props <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    prop.table(table(imp_data$person_dr))
  })

  # Create summary table
  dr_summary <- data.frame(
    Category = c("No_DR", "NPDR", "PDR"),
    Mean_Proportion = c(round(mean(dr_props[1,]), 3), 
                       round(mean(dr_props[2,]), 3),
                       round(mean(dr_props[3,]), 3)),
    SD = c(round(sd(dr_props[1,]), 3),
           round(sd(dr_props[2,]), 3), 
           round(sd(dr_props[3,]), 3))
  )
  kable(dr_summary, caption = "DR severity proportions across imputations")

  # Visual check if stripplot works with factors
  tryCatch({
    stripplot(mice_results, person_dr ~ .imp, pch = 20, cex = 1.2) |> print()
  }, error = function(e) {
    # Note: Stripplot not available for categorical person_dr
  })
}

# Check imputation quality for categorical variables
# Cataract surgery
if(sum(is.na(df_mice$catsurg_before_hitplus_cat)) > 0) {
  # Extract proportions for each imputation
  cat_props <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    prop.table(table(imp_data$catsurg_before_hitplus_cat))
  })

  # Create summary table for cataract surgery proportions
  if(is.matrix(cat_props)) {
    cat_summary <- data.frame(
      Level = rownames(cat_props),
      Mean_Proportion = apply(cat_props, 1, mean) |> round(3),
      SD = apply(cat_props, 1, sd) |> round(3)
    )
  } else {
    cat_summary <- data.frame(
      Level = "Cataract Surgery",
      Mean_Proportion = round(mean(cat_props), 3),
      SD = round(sd(cat_props), 3)
    )
  }
  kable(cat_summary, caption = "Cataract surgery proportions across imputations")
}
```

>>>>

# Initialize logging
log_file <- init_log("multiple_imputation")

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 50 imputations, 20 iterations (slow but thorough)
analysis_mode <- "test"  # Change to "final" for production run
log_message(paste0("Analysis mode: ", analysis_mode))

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 2
} else if (analysis_mode == "final") {
  n_imputations <- 50
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}

log_message(paste0("Imputations: ", n_imputations, ", Iterations: ", n_iterations))
```

@@

ps_model_vars <- variable_lists$ps_model_vars
outcome_model_vars <- variable_lists$outcome_model_vars

log_message(paste0("Loaded ", nrow(df_for_imputation), " observations"))
log_message(paste0("Unique patients: ", n_distinct(df_for_imputation$e_mrn_deidentified)))
```

@@

```{r mice-run}
log_message("Starting MICE imputation...")
imp_start <- Sys.time()

@@
